Some examples in extended logic counting classes

### 2016-12-05 LICS/horn-counting.tex
standarizing saluja

%!TEX root = main.tex

\subsection{Saluja Classes}

We define syntactically the classes $\XE{i}$ and $\XU{i}$ according to the following grammar:
\begin{enumerate}
	\item $\XE{0}$:
	\[
	E_0 ::= \varphi , \varphi \mbox{ is an {\sc FO}-extended quantifier-free $\L$-formula }.
	\]
	\item $\XU{0}$:
	\[U_0 ::= E_0.\]
	\item $\XE{i+1}$:
	\[	E_{i+1} ::= U_i \ \mid \ \exists x \, E_{i+1}.\]
	\item $\XU{i+1}$:
	\[U_{i+1} ::= E_i \ \mid \ \forall x \, U_{i+1}.\]
\end{enumerate}

A function $f$ is in $\XE{i}$ (resp. $\XU{i}$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

\begin{theorem} \label{fp1}
	$\XE{0} \subseteq$ {\sc FP}.
\end{theorem}

The {\em decision problem} associated to a function $f$ is defined by the language $L_f = \{\A \in \str \mid f(\A) > 0\}$.

\begin{theorem} \label{decisionptime}
	The decision problem associated to a function in $\E1$ is in \textsc{P}.
\end{theorem}

For a given pair of functions $f,g$, we define $f \dotminus g$ as follows:
\begin{eqnarray*}
	(f \dotminus g)(\A) =
	\begin{cases}
		f(\A)-g(\A), & \text{if }f(\A)>g(\A) \\
		0, & \text{if }f(\A) \leq g(\A).
	\end{cases}
\end{eqnarray*}
for every $\L$-structure $\A \in \str$. A function class $\F$ is {\em closed under substraction} if for every pair of functions $f,g \in \F$, it holds that $f \dotminus g \in \F$.

\begin{theorem} \label{sub-pnp}
	If $\E1$ is closed under substraction, then {\sc P} = {\sc NP}.
\end{theorem}

\begin{theorem} \label{sigma1strict}
	$\#\Sigma_1 \subsetneq \E1$
\end{theorem}

For a given function $f$, we define $f \dotminus 1$ as follows:
\begin{eqnarray*}
	f \dotminus 1(\A) =
	\begin{cases}
		f(\A)-1, & \text{if }f(\A) > 0 \\
		0, & \text{if }f(\A) = 0.
	\end{cases}
\end{eqnarray*}
for every $\L$-structure $\A \in \str$. A function class $\F$ is {\em closed under substraction by one} if for every function $f \in \F$, it holds that $f \dotminus 1 \in \F$.

\begin{theorem} \label{sigmafo-minusone}
	$\XE{1}$ is closed under substraction by one.
\end{theorem}

\begin{theorem} \label{dnf-pars}
	{\sc \#DNF} is hard for $\E1$ under parsimonious reductions. 
\end{theorem}

\begin{theorem} \label{nplusone-strict}
	$\U{1}$ with $n$ open first-order variables is properly contained in $\U{1}$ with $n+1$ open first-order variables for $n\in\N$.  
\end{theorem}

\subsection{Horn Counting Classes}

We define the syntactic classes $\XE{i}$ and $\XU{i}$ as follows. First we define extended Horn clauses.
\begin{eqnarray*}
	PL &::=& X_i(\x),\, i\in\N,\\
	NL &::=& \neg X_i(\x),\, i\in\N \ \mid \ \exists x\, NL,\\
	NC &::=& NL \ \mid \ \alpha, \alpha \mbox{ is an {\sc FO}-formula over } \L \ \mid \ (NC \vee NC),\\
	HC &::=& NC \ \mid \ (NC \vee PL) \ \mid \ PL,
\end{eqnarray*}
where $PL$ represents a positive literal, $NL$ is an \textit{extended} negative literal, $NC$ is an extended Horn clause, and $HC$ is an extended Horn formula. Now we define the syntax of the classes inductively.\textbf{}
\begin{enumerate}
	\item $\HE{0}$:
	\begin{eqnarray*}
		E_0 &::=& HC \ \mid \ E_0 \wedge E_0.
	\end{eqnarray*}
	\item $\HU{0}$:
	\begin{eqnarray*}
		U_0 &::=& E_0.
	\end{eqnarray*}
	\item $\HE{i+1}$:
	\begin{eqnarray*}
		E_{i+1} &::=& U_i \ \mid \ \exists x \, E_{i+1}.
	\end{eqnarray*}
	\item $\HU{i+1}$:
	\begin{eqnarray*}
		U_{i+1} &::=& E_i \ \mid \ \forall x \, U_{i+1}.
	\end{eqnarray*}
\end{enumerate}
A function $f$ is in $\XE{i}$ (resp. $\XU{i}$) if there is an $\L$-formula $\varphi$ defined by the grammar $E_i$ (resp. $U_i$) such that $f = f_{\varphi}$.

The class $\pe$ is defined in \cite{DBLP:conf/mfcs/PagourtzisZ06} as $\pe = \{f\mid f\in\#P \mbox{ and its decision version } L_{f}\in P\}$.

\begin{theorem} \label{sigma2-pe}
	$\HE{2} \subseteq \pe$
\end{theorem}

@@

\begin{theorem} \label{sigma2hard}
	$\shdhsat$ is hard for $\XE{2}$ under parsimonious reductions. 
\end{theorem}

>>>>

We define the fragments $\logeh{i}$ and $\loguh{i}$ as follows. First we define extended Horn clauses.
\begin{eqnarray*}
	PL &::=& X_i(\x),\, i\in\N,\\
	NL &::=& \neg X_i(\x),\, i\in\N \ \mid \ \exists x\, NL,\\
	NC &::=& NL \ \mid \ \alpha, \alpha \mbox{ is an {\sc FO}-formula} \ \mid \ (NC \vee NC),\\
	HC &::=& NC \ \mid \ (NC \vee PL) \ \mid \ PL,
\end{eqnarray*}
where $PL$ represents a positive literal, $NL$ is an \textit{extended} negative literal, $NC$ is an extended Horn clause, and $HC$ is an extended Horn formula. Now we define the syntax of the classes inductively.
\begin{align*}
\logeh{0} &::= HC \ \mid \ \logeh{0} \wedge \logeh{0}, \\
\loguh{0} &::= \logeh{0}, \\ 
\logeh{i+1} &::= \loguh{i} \ \mid \ \exists x\, \logeh{i+1}, \\
\loguh{i+1} &::= \logeh{i} \ \mid \ \forall x\, \loguh{i+1}.
\end{align*}
\begin{theorem} \label{sigma2-pe}
	$\eqso(\loguh{2}) \subseteq \pe$
\end{theorem}

@@

\begin{theorem} \label{sigma2hard}
	$\shdhsat$ is hard for $\eqso(\loguh{2})$ under parsimonious reductions. 
\end{theorem}

### 2016-12-05 LICS/preliminaries.tex
standarizing saluja

Here come the preliminaries

>>>>

A $\mathbb{N}$ relational signature $\R$ is a finite set $\{R_1, \ldots, R_k\}$, where each $R_i$ ($1 \leq i \leq k$) is a relation name with an associate arity greater than 0, which is denoted by $\arity(R_i)$. A finite structure over $\R$ (or just finite $\R$-structure) is a tuple $\A = \langle A, R_1^\A, \ldots, R_k^\A \rangle$ such that $A$ is a finite set and $R_i^\A \subseteq A^{\arity(R_i)}$ for every $i \in \{1, \ldots, k\}$. A finite $\R$-structure $\A$ is said to be ordered if $<$ is a binary predicate name in $\R$ and $<^\A$ is a linear order on $A$. Let $\str[\R]$ be the class of all finite $\R$-structures and $\ostr[\R]$ be the class of all ordered finite $\R$-structures.

From now on, assume given disjoint infinite sets $\fv$ and $\sv$ of first-order variables and second-order variables, respectively. Notice that every variable in $\sv$ has an associated arity, which is denoted by $\arity(X)$. Then given a relational signature $\R$, the set of second-order logic formulas ($\so$-formulas) over $\R$ is given by the following grammar:
\begin{eqnarray*}\ 
	\varphi &:=& R(\bar u) \ \mid\  
	X(\bar v)  \ \mid\ 
	\neg \varphi \ \mid\ 
	(\varphi \vee \varphi) \ \mid\ 
	\exists x \, \varphi \ \mid\ 
	\exists X \, \varphi
\end{eqnarray*}
where $R \in \R$, $\bar u$ is a tuple of (non-necessarily distinct) variables from $\fv$ whose length is $\arity(R)$, $X \in \sv$, $\bar v$ is a tuple of (non-necessarily distinct) variables from $\fv$ whose length is $\arity(X)$, and $x \in \fv$.

To define the semantics of $\so$, we need to introduce some terminology. Given a relational signature $\R$ and a finite $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ is a total function from $\fv$ to $A$, while a second-order assignment $V$ for $\A$ is a total function with domain $\sv$ that maps each $X \in \sv$ to a subset of $A^{\arity(X)}$. Moreover, given a first-order assignment $v$ for $\A$, $x \in \fv$ and $a \in A$, we denote by $v[a/x]$ a first-order assignment such that $v[a/x](x) = a$ and $v[a/x](y) = v(y)$ for every $y \in \fv$ distinct from $x$. Similarly, given a second-order assignment $V$ for $\A$, $X \in \sv$ and $B  \subseteq A^{\arity(X)}$, we denote by $V[B/X]$ a second-order assignment such that $V[B/X](X) = B$ and $V[B/X](Y) = V(Y)$ for every $Y \in \sv$ distinct from $X$.

Assume that $\varphi$ is an $\so$-formula over a signature $\R$. Then given a finite $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ and a second-order assignment $V$ for $\A$, we say that $(\A, v, V)$ satisfies $\varphi$, denoted by $(\A, v, V) \models \varphi$, if: (1) $\varphi$ is the formula $R(x_1, \ldots, x_\ell)$ and $(v(x_1), \ldots, v(x_\ell)) \in R^\A$; (2) $\varphi$ is the formula $X(x_1, \ldots, x_m)$ and $(v(x_1), \ldots, v(x_m)) \in V(X)$; (3) $\varphi$ is the formula $\neg \psi$ and it not the case that $(\A, v, V) \models \psi$; (4) $\varphi$ is the formula $(\varphi_1 \vee \varphi_2)$, and $(\A, v, V) \models \varphi_1$ or $(\A, v, V) \models \varphi_2$; (5) $\varphi$ is the formula $\exists x \, \psi$ and there exists $a \in A$ such that $(\A, v[a/x], V) \models \psi$; or (6) $\varphi$ is the formula $\exists X \, \psi$ and there exists $B \subseteq A^{\arity(X)}$ such that $(\A, v, V[B/X]) \models \psi$.  As usual, we consider the propositional operators $\wedge$, $\rightarrow$, and $\leftrightarrow$ that can be obtained from $\vee$ and $\neg$. 
%Moreover, we use the abbreviations $x \not \leq y$ and $x \notin X$ for the negation of the atoms $\leq$ and $\in$. 
%Finally, we consider standard abbreviations of formulas that can be defined in $\so$-logic (actually, $\fo$-logic) like $\first(x) := \fa{x} y \leq x$ and $\last(x) := \fa{x} x \leq y$ to denote the first and last element of the linear order $\leq$, respectively, and $\succesor(x,y) := x \leq y \wedge y \not\leq x \wedge \fa{z} ( z \leq x \vee y \leq z)$ to denote the successor relation.

\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is a finite $\R$-structure with a domain $A$ containing $n$ elements, and assume that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a set $\bbD$ and a relational signature $\R$, a function $f$ is said to be from $\R$ to $\bbD$ if $f$ is a total function from $\{\enc(\A) \mid \A \in \str[\R]\}$ to $\bbD$. Moreover, a set $\CC$ of functions over $\bbD$ is said to be a {\em function complexity class over $\bbD$} if every $f \in \CC$ is a function from some relational signature $\R$ to $\bbD$.

If $\KK$ is a class of finite structures and $f$ is a function from a relational signature $\R$ to $\bbD$, we denote by $\res{f}{\KK}$ the restriction of $f$ to $\KK$, that is, a function such that: the domain of $\res{f}{\KK}$ is $\{ \enc(\A) \mid \A \in \str[\R] \cap \KK\}$, and 
$\res{f}{\KK}(\enc(\A)) = f(\enc(\A))$ for every $\A \in \str[\R] \cap \KK$.
